Seeing Double: Transient Dips and Photometric Binarity in PTFO$\,$8-8695

### 2020-04-21 paper/ms.tex
fix TeX column width

\begin{abstract}
  PTFO$\,$8-8695b is a candidate hot Jupiter in the 7--10 million year
  old Orion-OB1a cluster. We inspected data from TESS and Gaia to
  clarify whether it is truly a planet.  The Gaia data show that
  PTFO$\,$8-8695 is a photometric binary with respect to members of
  its kinematic group.  The TESS lightcurve shows that the dominant
  variability in this system is a sinusoid with a ``long'' period
  $P_{\rm \ell}=11.96\,$hr, presumably caused by stellar rotation.
  Also present is a complex signal, previously identified as the
  planet candidate, that repeats with a ``short'' period $P_{\rm s}=
  10.74\,$hr.  The two signals beat every 4.48 days.  Although there
  is a dip in the short-period signal, ground-based photometry from
  the past decade shows that the orbital phase of the dip seems to
  have instantaneously jumped, at least once, and perhaps twice.
  Planets do not ``jump'' in orbital phase.  Given the evidence, we
  believe that PTFO$\,$8-8695 is a binary M dwarf in which one star
  shows the long rotation signal, and the other star shows
  ``transient dipping'' that has been seen in a few other young
  weak-lined T Tauri stars.  The origin of these transient dips is
  still undetermined, but our preferred explanation is eclipses of
  clouds of gas or dust at the Keplerian co-rotation radius.
\end{abstract}

@@

\section{Introduction}
If \ptfob\ were a planet, it would be exceptional.  Transiting a
sub-10$\,$Myr old weak-lined T Tauri M dwarf in Orion, it would be the
youngest hot Jupiter known \citep{van_eyken_ptf_2012}.  Its orbital
period of only 10.7 hours would also give it the shortest period of
any known hot Jupiter.  With such a short period, it would almost
certainly have filled its Roche lobe, and would be actively losing
mass to its host star.  Not only that, but the rapidly rotating host
star would also be oblate and gravity darkened, and so the planet's
orbit would likely precess into and out of transitability
\citep{barnes_measurement_2013,ciardi_followup_2015,kamiaka_revisiting_2015}.

@@

Another first could be the direct detection of H$\alpha$ emission from
the planet itself \citep{johnskrull_h_2016}.  While the stellar
chromosphere emits in H$\alpha$, it seems that there is an additional
excess H$\alpha$ emission that could be in phase with the planetary
orbit.  The average velocity width of the excess H$\alpha$ emission is
87$\,$km$\,$s$^{-1}$, and its equivalent width is 70-80\% that of the
stellar chromosphere \citep{johnskrull_h_2016}.  The proposed
explanation is that outflowing mass from the planet may explain this
excess emission as well \citep{johnskrull_h_2016}.

There are perhaps a few challenges to the planetary interpretation (if
these ``features'' are not already seen as such).  They include that
the planet does not seem to emit infrared radiation in occulation, at
least anywhere near the expected amplitude \citep{yu_tests_2015}.  In
addition, despite measurement attempts by multiple investigators, it
does not seem to show the Rossiter effect at the amplitude expected
given the rapid stellar rotation and large planet size
\citep{yu_tests_2015,ciardi_followup_2015}.  Finally, detailed
modelling of the ``precession + gravity darkening'' transits has shown
that the necessary degree of gravity darkening is too great, given the
spectroscopically observed equatorial velocity
\citep{howarth_reappraisal_2016}.  Additionally, as the
gravity-darkened star precessed about its rotation axis, it would need
to show photometric variability that has not been observed.  While the
planetary interpetation clearly faces challenges, alternative
explanations do as well.  High-latitude accretion hotspots might
produce the observed H$\alpha$ variability, and require fine-tuning
produce dips of the approach duration. Furthermore, \ptfo\ does not
have an infrared (IR) excess associated with the presence of an inner
disk \citep[{\it e.g.},][Figure~18]{yu_tests_2015}.  Low-latitude
starspots, hot or cold, struggle to produce the necessary dip
durations.

Alternatively, between 0.1\% and 1\% of rapidly rotating low-mass
stars in $\mathcal{O}$(10)$\,$Myr old associations show narrow dips in
phase with strong stellar rotation signals \citep{rebull_usco_2018}.
The dips can persist over months, but their depths often vary, and
sometimes change immediately after stellar flares.  The explanation
proposed by \citet{stauffer_orbiting_2017} and
\citet{david_transient_2017} is that a circumstellar cloud of dust or
gas might be orbiting near the Keplerian co-rotation radius.  For this
explanation to be viable for \ptfo, a clear determination of the
stellar rotation period is necessary.  To date this has not been
possible
\citep{van_eyken_ptf_2012,koen_multicolour_2015,raetz_yeti_2016}.

@@

Gaia also provided astrometric parameters for tens of thousands of
young stars in the Orion complex.  The stellar populations in giant
molecular cloud complexes are not monolithic; substructured groups are
the norm \citep{briceno_lowmassOB_2007}.  The Orion molecular cloud
complex in particular has numerous subgroups, with ages spanning 0.5
to 15$\,$Myr. For an incomplete sampling, see for instance
\citet{briceno_cida_2005,jeffries_kinematic_2006,briceno_25_2007,kounkel_apogee2_2018}
and \citet{briceno_cidaII_2019}.

@@

With the Gaia astrometry, it has become clear that 25$\,$Ori itself
has further subgroups
\citep{kounkel_apogee2_2018,briceno_cidaII_2019}.  In describing the
cluster membership of \ptfo, we follow the notation and results of
\citet{kounkel_apogee2_2018}.  These authors combined astrometric data
from Gaia DR2 with near-infrared spectra from APOGEE-2
\citep{gunn_sdss_2006,majewski_apache_2017,blanton_sloan_2017,zasowski_target_2017,cottle_apogee2_2018}.
They performed a hierarchical clustering on the six dimensional
position and velocity information to identify subgroups within the
Orion complex.  From smallest to largest, \ptfo\ was identified
as being a member of the following hierarchical subgroups:
\begin{equation}
  %{\rm PTFO\,8\text{-}8695}
  % \in
  {\rm 25\,Ori\text{-}1}
  \subset {\rm 25\,Ori}
  \subset {\rm Orion\ OB1a}
  \subset {\rm Orion\ D},
\end{equation}
where from set-notation, `$\subset$' denotes ``is a proper subset of''.
25$\,$Ori-1 is the largest subgroup of 25$\,$Ori, with 149 identified
members.
Its mean age from its CMD was found to be 6.9$\,$Myr, and from its HR
diagram 8.5$\,$Myr \citep{kounkel_apogee2_2018}.
\citet{kounkel_apogee2_2018} identified seven other smaller groups in
the Orion complex near the Be star 25$\,$Ori. These groups received
higher numbers, {\it e.g.}, 25$\,$Ori-2 (${\rm Age}_{\rm CMD} =15.1\,$Myr; ${\rm Age}_{\rm CMD} =12.9\,$Myr; see also \citealt{briceno_cidaII_2019}).

@@

The implication is that either {\it (i)} \ptfo\ is notably younger
than the kinematically identical 25$\,$Ori-1 members, or {\it (ii)} \ptfo\
is a photometric binary.
Since there is no a priori reason to suspect an age difference,
but we have resolved two
separate photometric signals, we favor the binary interpretation.

@@

Figure~\ref{fig:o_minus_c} shows two additional strange features:
{\it (i)}  multiple dips per cycle, and {\it (ii)} a set of dips numerically coincident
with phase $(P_{\rm \ell} - P_{\rm s}) / P_{\rm s}$.
The observation of multiple
dips per cycle in 2015 seems to have been detected by both \citet{yu_tests_2015}
and \citet{tanimoto_evidence_2020}.
It therefore seems credible.
Inspecting the \citet{tanimoto_evidence_2020} lightcurves,
the claim of multiple dips per cycle in December 2018 at phase 0 and -0{.}47
seems much less plausible---the phase -0{.}47 dips are strongly detected,
while the suggested phase 0 dip is not clearly present.

We are not sure what to make of the numerical coincidence.
The ratio of long to short periods is roughly 10:9, which does produce a beat.
However it is not clear that this would obviously translate into
an observational bias unless by some fluke three season's worth 
of observations managed to only observe every ninth dip.
This is of course not the case, and we therefore leave this 
curiosity as observation {\it sans} interpretation.

@@

\subsection{Short period out-of-dip modulation}
If there were a giant planet transiting \ptfo, it would tidally
distort the host star, and cause ellipsoidal photometric modulations.
The amplitude of the ellipsoidal distortion for a $1\,M_{\rm Jup}$
companion would be about 1400$\,$ppm
\citep{shporer_astrophysics_2017}.  This is significantly larger than
the typical ellipsopidal modulation induced by close-in giant planets
because the host star is puffy, and still on the pre-main-sequence.
For our estimate, we assumed $R_\star = 1.39 R_\odot$, and $M_\star =
0.39 M_\odot$ \citep{van_eyken_ptf_2012}.

@@

>>>>

\begin{abstract}
  PTFO$\,$8-8695b is a candidate hot Jupiter in the 7--10 million year
  old Orion-OB1a cluster. We inspected data from TESS and Gaia to
  clarify whether it is truly a planet.  The Gaia data show that
  PTFO$\,$8-8695 is a photometric binary with respect to members of
  its kinematic group.  The TESS lightcurve shows that the dominant
  variability in this system is a sinusoid with a long period
  $P_{\rm \ell}=11.96\,$hr, presumably caused by stellar rotation.
  Also present is a complex signal, previously identified as the
  planet candidate, that repeats with a short period $P_{\rm s}=
  10.74\,$hr.  The two signals beat every 4.48 days.  Although there
  is a dip in the short-period signal, ground-based photometry from
  the past decade indicates that the orbital phase of the dip seems to
  have instantaneously jumped, at least once, and perhaps twice.
  Planets do not ``jump'' in orbital phase.  Given the available
  evidence, we believe that PTFO$\,$8-8695 is a binary M dwarf in
  which one star shows the long rotation signal, and the other star
  shows ``transient dipping'' that has been seen in a few other young
  weak-lined T Tauri stars.  The origin of these transient dips is
  still undetermined, but our preferred explanation is eclipsing
  clouds of gas or dust at the Keplerian co-rotation radius.
\end{abstract}

@@

\section{Introduction}
If \ptfob\ were a planet, it would be exceptional.  At below 10
million years of age, it could be the youngest hot Jupiter known
\citep{van_eyken_ptf_2012}.  Its orbital period of only 10.7 hours
around a weak-lined T Tauri M dwarf would also give it the shortest
period of any known hot Jupiter.  With such a short period, it would
almost certainly have filled its Roche lobe, and would be actively
losing mass to its host star.  Not only that, but the rapidly rotating
host star would also be oblate and gravity darkened, and so the
planet's orbit would likely precess into and out of transitability
\citep{barnes_measurement_2013,ciardi_followup_2015,kamiaka_revisiting_2015}.

@@

Another first could be the direct detection of H$\alpha$ emission from
the planet itself \citep{johnskrull_h_2016}.  While the stellar
chromosphere emits in H$\alpha$, it seems that there is an additional
excess H$\alpha$ emission that could be in phase with the planetary
orbit.  The average velocity width of the excess H$\alpha$ emission is
87$\,$km$\,$s$^{-1}$, and its equivalent width is 70-80\% that of the
stellar chromosphere \citep{johnskrull_h_2016}.  The proposed
explanation is that outflowing mass from the planet could explain this
excess emission as well \citep{johnskrull_h_2016}.

There are perhaps a few challenges to the planetary interpretation (if
these ``features'' are not already seen as such).  They include that
the planet does not seem to emit infrared radiation in occulation, at
least at the expected amplitude \citep{yu_tests_2015}.  In
addition, despite measurement attempts by multiple investigators, it
does not seem to show the Rossiter effect at the amplitude expected
given the rapid stellar rotation and large planet size
\citep{yu_tests_2015,ciardi_followup_2015}.  Finally, detailed
modelling of the ``precession + gravity darkening'' transits has shown
that the necessary degree of gravity darkening is too great, given the
spectroscopically observed equatorial velocity
\citep{howarth_reappraisal_2016}.  Additionally, as the
gravity-darkened star precessed about its rotation axis, it would need
to show photometric variability that has not been observed.

While the planetary interpetation clearly faces challenges,
alternative explanations do as well.  High-latitude accretion hotspots
might produce the observed H$\alpha$ variability, and require
fine-tuning produce dips of the approach duration. Furthermore, \ptfo\
does not have an infrared (IR) excess associated with the presence of
an inner disk \citep[{\it e.g.},][Figure~18]{yu_tests_2015}.
Low-latitude starspots, hot or cold, struggle to produce the necessary
dip durations.

Another potential explanation is that between 0.1\% and 1\% of rapidly
rotating low-mass stars in $\mathcal{O}$(10)$\,$Myr old associations
show narrow dips in phase with strong stellar rotation signals
\citep{rebull_usco_2018}.  The dips can persist over months, but their
depths often vary, and sometimes change immediately after stellar
flares.  The explanation proposed by \citet{stauffer_orbiting_2017}
and \citet{david_transient_2017} to explain this novel class of
variable stars is that a circumstellar cloud of dust or gas might be
orbiting near the Keplerian co-rotation radius.  Given available
ground-based data, this explanation is not clearly viable for \ptfo,
because a clear determination of the stellar rotation period has not
yet been possible
\citep{van_eyken_ptf_2012,koen_multicolour_2015,raetz_yeti_2016}.

@@

Gaia also provided astrometric parameters for tens of thousands of
young stars in the Orion complex.  Stellar populations in giant
molecular cloud complexes are not monolithic; substructured groups are
the norm \citep{briceno_lowmassOB_2007}.  The Orion molecular cloud
complex in particular has numerous subgroups, with ages spanning 0.5
to 15$\,$Myr. For an incomplete sampling, see for instance
\citet{briceno_cida_2005,jeffries_kinematic_2006,briceno_25_2007,kounkel_apogee2_2018}
and \citet{briceno_cidaII_2019}.

@@

With the Gaia astrometry, it has become clear that 25$\,$Ori itself
has further subgroups
\citep{kounkel_apogee2_2018,briceno_cidaII_2019}.  In describing the
cluster membership of \ptfo, we follow the notation and results of
\citet{kounkel_apogee2_2018}.  These authors combined astrometric data
from Gaia DR2 with near-infrared spectra from APOGEE-2
\citep{gunn_sdss_2006,majewski_apache_2017,blanton_sloan_2017,zasowski_target_2017,cottle_apogee2_2018}.
They performed a hierarchical clustering on the six dimensional
position and velocity information to identify subgroups within the
Orion complex.  From smallest to largest, \ptfo\ was identified
as a member of the following hierarchical subgroups:
\begin{equation}
  %{\rm PTFO\,8\text{-}8695}
  % \in
  {\rm 25\,Ori\text{-}1}
  \subset {\rm 25\,Ori}
  \subset {\rm Orion\ OB1a}
  \subset {\rm Orion\ D},
\end{equation}
where from set-notation, `$\subset$' denotes ``is a proper subset
of''.  25$\,$Ori-1 is the largest subgroup of 25$\,$Ori, with 149
identified members.  Its mean age from its CMD was found to be
6.9$\,$Myr, and from its HR diagram 8.5$\,$Myr
\citep{kounkel_apogee2_2018}.  \citet{kounkel_apogee2_2018} identified
seven other smaller groups in the Orion complex near the Be star
25$\,$Ori. These groups received higher numbers, {\it e.g.},
25$\,$Ori-2 (${\rm Age}_{\rm CMD} =15.1\,$Myr; ${\rm Age}_{\rm CMD}
=12.9\,$Myr; see also \citealt{briceno_cidaII_2019}).

@@

The implication is that either {\it (i)} \ptfo\ is notably younger
than the kinematically identical 25$\,$Ori-1 members, or {\it (ii)}
\ptfo\ is a photometric binary.  Since there is no a priori reason to
suspect an age difference, but we have resolved two separate
photometric signals, the binary interpretation seems somewhat more
likely.

@@

Figure~\ref{fig:o_minus_c} shows two additional strange features: {\it
(i)}  multiple dips per cycle, and {\it (ii)} a set of dips
numerically coincident with phase $(P_{\rm \ell} - P_{\rm s}) / P_{\rm
s}$.  The observation of multiple dips per cycle in 2015 was seen
independently by both \citet{yu_tests_2015} and
\citet{tanimoto_evidence_2020}.  It therefore seems credible.
Inspecting the \citet{tanimoto_evidence_2020} lightcurves, the claim
of multiple dips per cycle in December 2018 at phase 0 and -0{.}47
seems less plausible---the phase -0{.}47 dips are strongly detected,
while the suggested phase 0 dip is not clearly present in the data.

We are not sure what to make of the numerical coincidence.  The ratio
of long to short periods is roughly 10:9.  It is not clear that this
would obviously translate into an observational bias unless by some
fluke three season's worth of observations managed to only observe
every ninth dip.  This is of course not the case, and we therefore
leave this curiosity as observation {\it sans} interpretation.

@@

\subsection{Short period out-of-dip modulation}

%FIXME write it from here!

If there were a giant planet transiting \ptfo, it would tidally
distort the host star, and cause ellipsoidal photometric modulations.
The amplitude of the ellipsoidal distortion for a $1\,M_{\rm Jup}$
companion would be about 1400$\,$ppm
\citep{shporer_astrophysics_2017}.  This is significantly larger than
the typical ellipsopidal modulation induced by close-in giant planets
because the host star is puffy, and still on the pre-main-sequence.
For our estimate, we assumed $R_\star = 1.39 R_\odot$, and $M_\star =
0.39 M_\odot$ \citep{van_eyken_ptf_2012}.

@@

\subsection{Similar Lightcurves}

Figure~\ref{fig:brethren}
